Prestige bias in cultural evolutionary dynamics

### 2023-01-05 ms/ms.tex
revise with SE

% Title page
\title{Cultural evolutionary dynamics with success and influence bias}
% Authors
\renewcommand\Affilfont{\small}

@@

\citet[Ch.~5]{evolutionBook} have suggested that the evaluation of success can be divided into three groups: \emph{direct bias}, \emph{indirect bias} and \emph{frequency-dependent bias}.
A direct bias occurs when one phenotype is more attractive than other phenotypes, and is evaluated by \emph{directly} testing the trait.
For example, an individual observing a ping-pong match can attempt the observed paddle grips to determine which grip is better.
A frequency-dependent bias occurs when the probability to copy a phenotype is higher or lower than the frequency of the phenotype among demonstrators. 
For example, suppose the common paddle grip is used by 60\% of the demonstrators; if the this grip is adopted by 80\% of copiers, then transmission is under positive frequency bias, also called \emph{conformity}; if it is adopted by 40\% of copiers, then transmission is under negative frequency bias, or \emph{non-conformity} \citep{conformism}. 
The effects of conformity and non-conformity on cultural evolution have been studied with both models \citep{anticonformity,Denton2021,conformity} and experiments \citep{negativeFrequency}.
An indirect bias occurs when an individual uses one phenotype to determine the attractiveness of another phenotype.
For example, an observer may copy the paddle grip of the ping-pong player who scored more points in the match, thus indirectly evaluating the grip by the points scored.
This, however, may cause mismatches between the copied trait and the rest of the cultural or genetic repertoire of the individual \citep{Kolodny2022}.
Furthermore, \citet[Ch. 8]{evolutionBook} have suggested that maladaptive traits may spread widely in a population if indirect biases are strong enough by a runaway process caused by a cultural equivalent of sexual selection \citep{sexualSelectionBook}.
For example, \citet{cooperation} have shown that helping behaviors can evolve due to horizontal transmission bias even without any benefit to the recipient, or when the benefit is much larger than the cost.

@@

\subsection*{Dichotomous trait}
To allow for mathematical analysis of the model, we introduce a simplified version where the trait only has two phenotypes: the optimal phenotype $\hat{A}$ and the sub-optimal phenotype $A$. 
All role-models with the same phenotype will contribute to the probability of that phenotype to be transmitted, and thus influence is determined by the number of copiers that have already chosen a role-model with either phenotype.
In addition, for simplicity and for easier mathematical analysis, we assume $\alpha$ is homogeneous, such that we have exchangeability between role-models.
Therefore, the probability of the $i$-th copier to copy phenotype $A$ is
\begin{equation}\label{eq:binary-model}
G_{i,A} = \frac{(N-C)\alpha'\beta(A) + K_{i,A}}{i-1 + (N-C)\alpha'\beta(A) + C\alpha'\beta(\hat{A})} = \frac{(N-C)\alpha'\beta(A) + K_{i,A}}{i-1 + (N-C)\alpha'\beta(A) + \alpha'C} \;,
\end{equation}
where $C$ is the number of role-models with trait $\hat{A}$ and $K_{i,A}$ is the number of copiers that already chose $A$ when copier $i$ chooses a role-model. $\alpha'$ is the odd ratio of the bias weight $\alpha'= \frac{\alpha}{1-\alpha}$.
We prove this equation later, based on the proof of the Dirichlet-Multinomial approximation. % TODO what do you mean "we prove this later"..?
The rest of the details follow the continuous trait model.

@@

Using our effective selection coefficient, $1-\beta$, and effective population size, $N_e$, with the population-genetics fixation probability approximation given by \citet[eq.~8]{kimura}, we get the following result:
\begin{result}[Fixation probability]
The fixation probability is approximately
\begin{equation}\label{eq:kimura_p}
\pi = \frac{1-e^{-2(1-\beta)N_e x}}{1-e^{-2(1-\beta)N_e}}
\end{equation}
where $x$ is the initial frequency of the favored phenotype $\hat{A}$.
\end{result}

@@

\begin{result}[Fixation time] The fixation time (conditioned on fixation) is approximately
\begin{equation}
T = J_1 + \frac{1-u(x)}{u(x)}\cdot J_2,
\end{equation}
where, given $S=N_e(1-\beta)$ and $x=\frac{1}{N_e}$,
\begin{equation}
u(x)=\frac{1-e^{-2Sx}}{1-e^{2S}}
\end{equation}
and,
\begin{equation}
J_1 = \frac{2}{(1-\beta)(1-e^{-2S})}\int_x^1 \frac{(e^{2S\xi}-1)(e^{-2S\xi}-e^{-2S})}{\xi(1-\xi)}d\xi,
\end{equation}
\begin{equation}
J_2= \frac{2}{(1-\beta)(1-e^{-2S})}\int_0^p \frac{(e^{2S\xi}-1)(1-e^{-2S\xi})}{\xi(1-\xi)}d\xi
\end{equation}

\begin{equation}\label{eq:kimura_t}
T = \frac{1-\pi}{1-\beta}\int_0^x\frac{e^{2(1-\beta) \xi}-1}{\xi(1-\xi)}d\xi+ \frac{\pi}{1-\beta}\int_x^1\frac{1-e^{-2(1-\beta)(1-\xi)}}{\xi(1-\xi)}d\xi
\end{equation}
\end{result}
Note that these integrals cannot be solved in closed form, so we can only estimate them numerically.

\paragraph{Numerical validation.}
We compare our approximations (\cref{eq:kimura_p,eq:kimura_t}) with results of simulations of our dichotomous model using various $\alpha$ and $\beta$ values, as well as simulations of the Wright-Fisher model, using the effective selection coefficient, $s_e=1-\beta$, and effective population size, $N_e=\alpha N + (1-\alpha)$. 
We find see that the two models have similar dynamics, and both are well approximated our approximations (\Cref{fig:var_alpha}).

\begin{figure}[h]
    \includegraphics[width=\linewidth]{../figures/final/kimura_var.pdf}
  \caption{\textbf{Fixation probability and time in a constant environment.}
  Fixation probability and time (in generations) as a function of the success-bias weight $\alpha$ (bottom x-axis), or effective population size $N_e$ (top x-axis) in the top row, and as a function of the success coefficient, $1-\beta$, on the bottom row.
  The approximation (black; \cref{eq:kimura_p}) agrees with both DM simulations (green) and Wright-Fisher simulation (orange).
  Fixation probability (A) is bounded by $2s_e=2(1-\beta)$ (blue).
  Markers are averages of $10,000$ simulations, error bars show 95\% confidence intervals for panels A and B and 75\% for panels C and D.
   Here, Population size, $N=1,000$; phenotype values, $\hat{A}=1$, $A=0.7$ (A and B), $A = a \cdot \hat{A}$ with $0.01 \le a \le 0.99$ (panels C and D); success coefficient, $1-\beta=s=0.044$ (A and B); success-bias weight, $\alpha=0.01$ (panels C and D).}
  \label{fig:var_alpha}
\end{figure}

@@

Note that here, we have the average selection coefficient during a cycle of $n$ generations as the selection coefficient \cref{eq:kimura_p}.
Using the drift and diffusion terms and following \citet{changeEnv}, we can approximate the fixation probability in a changing environment using
\begin{equation}\label{eq:ch_env}
\tilde\pi = \frac{1-e^{-2 \frac{S_n}{n} N_e x}}{1-e^{-2 \frac{S_n}{n} N_e}} \;.
\end{equation}

\paragraph{Numerical validation.}
Comparing our approximation (\cref{eq:ch_env}) to simulations, we find that the approximation fits simulation results well for variable bias weights, $\alpha$, which corresponds to the effective population size (\Cref{fig:ch_env_alpha_beta}A).

@@

We found that for a large $k/l$ ratio (with a constant cycle length, $n=k+l=100$), the changing environment approximation (\cref{eq:ch_env}) converges to the constant environment approximation (\cref{eq:kimura_p}), see \Cref{fig:ch_env_alpha_beta}C and \Cref{fig:ch_env_alpha_beta}D. \\

% TODO the next paragraph is strange...
\begin{result}
The effective population size doesn't affect the goodness of fit of the approximation. The success coefficient however, must be even lower than before for the approximation to fit.
\end{result}
The proof is seen in \Cref{fig:ch_env_alpha_beta}A,B. The approximation follows the trend of the simulation results for all $\alpha$ values.
When increasing the success coefficient to more than 0.15, the simulation results were located above the changing environment approximation, and below the constant environment approximation. We believe the reason is the structure of the cycle.
Our proof and approximation in the changing environment are for a large amount of cycles, and when the success coefficient is too high, there might be very few cycles. Either the ideal trait is copied by enough copiers so that the influence is sufficient to negate the success bias when the cycle changes (and the trait favored by the bias becomes the disfavored), or the opposite happens, and the ideal trait gets extinct before there were enough copiers that copied it.
We then tried to change the ratio between the number of cycles where $\hat{A}$ is favored and disfavored. We showed that the approximation fits well regardless of the ratio, but when the ratio of favored generations to disfavored ones is very high, it is very similar to a constant environment model.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Discussion}
During cultural transmission, cultural traits such as attitudes, values, beliefs, and behavioral patterns are transmitted between individuals, for example via copying and social learning.
Some cultural traits or cultural role-models may be copied more often due to transmission biases. 
A common bias is success bias, in which copiers are more likely to copy a successful role-model. Many models assume that success can be accurately estimated.
However, it has been suggested that because it is hard to estimate success, a more common bias is \emph{prestige bias}---a bias towards role-models perceived to be successful in other traits \citep{fijian_social_bias}.

>>>>

% Title page
\title{Prestige bias in cultural evolutionary dynamics}
% Authors
\renewcommand\Affilfont{\small}

@@

% TODO edit so that we focus on indirect success - aka prestige - and decompose it to content based - which we call success - and context based - which we call influence
\citet[Ch.~5]{evolutionBook} have suggested that the evaluation of success can be divided into three groups: \emph{direct bias}, \emph{indirect bias} and \emph{frequency-dependent bias}.
A direct bias occurs when one phenotype is more attractive than other phenotypes, and is evaluated by \emph{directly} testing the trait.
For example, an individual observing a ping-pong match can attempt the observed paddle grips to determine which grip is better.
A frequency-dependent bias occurs when the probability to copy a phenotype is higher or lower than the frequency of the phenotype among demonstrators. 
For example, suppose the common paddle grip is used by 60\% of the demonstrators; if the this grip is adopted by 80\% of copiers, then transmission is under positive frequency bias, also called \emph{conformity}; if it is adopted by 40\% of copiers, then transmission is under negative frequency bias, or \emph{non-conformity} \citep{conformism}. 
The effects of conformity and non-conformity on cultural evolution have been studied with both models \citep{anticonformity,Denton2021,conformity} and experiments \citep{negativeFrequency}.
An indirect bias occurs when an individual uses one phenotype to determine the attractiveness of another phenotype.
For example, an observer may copy the paddle grip of the ping-pong player who scored more points in the match, thus indirectly evaluating the grip by the points scored.
This, however, may cause mismatches between the copied trait and the rest of the cultural or genetic repertoire of the individual \citep{Kolodny2022}.
Furthermore, \citet[Ch. 8]{evolutionBook} have suggested that maladaptive traits may spread widely in a population if indirect biases are strong enough by a runaway process caused by a cultural equivalent of sexual selection \citep{sexualSelectionBook}.
For example, \citet{cooperation} have shown that helping behaviors can evolve due to horizontal transmission bias even without any benefit to the recipient, or when the benefit is much larger than the cost.

@@

\subsection*{Dichotomous trait}
To allow for mathematical analysis of the model, we introduce a simplified version where the trait only has two phenotypes: the optimal phenotype $\hat{A}$ and the sub-optimal phenotype $A$. 
All role-models with the same phenotype will contribute to the probability of that phenotype to be transmitted, and thus influence is determined by the number of copiers that have already chosen a role-model with either phenotype.
In addition, for simplicity and for easier mathematical analysis, we assume $\alpha$ is homogeneous, such that we have exchangeability between role-models.
Therefore, the probability of the $i$-th copier to copy phenotype $A$ is
\begin{equation}\label{eq:binary-model}
G_{i,A} = \frac{(N-X)\alpha'\beta(A) + K_{i,A}}{i-1 + (N-X)\alpha'\beta(A) + \alpha'X} \;,
\end{equation}
where $X$ is the number of role-models with trait $\hat{A}$, $K_{i,A}$ is the number of copiers that already chose $A$ when copier $i$ chooses a role-model, $\beta(\hat{A})=1$, and $\alpha'= \frac{\alpha}{1-\alpha}$ is the odd ratio of the bias weight. Complementing this, the probability of the $i$-th copier to copy phenotype $\hat{A}$ is $G_{i, \hat{A}} = 1-G_{i,A}$.
The rest of the details follow the continuous trait model.

@@

Using our effective selection coefficient, $1-\beta$, and effective population size, $N_e$, with the population-genetics fixation probability approximation given by \citet[eq.~8]{kimura}, we get the following result:
\begin{result}[Fixation probability]
The fixation probability is approximately
\begin{equation}\label{eq:kimura_p}
\pi(x) = \frac{1-e^{-2(1-\beta)N_e x}}{1-e^{-2(1-\beta)N_e}}
\end{equation}
where $x$ is the initial frequency of the favored phenotype $\hat{A}$.
\end{result}

@@

% TODO SAAR check these equations vs what you have in the source code
\begin{result}[Fixation time] The fixation time (conditioned on fixation) from an initial frequency $x$ is approximately
\begin{equation} \label{eq:kimura_t}
T(x) = J_1(x) + \frac{1-\pi(x)}{\pi(x)}\cdot J_2(x),
\end{equation}
where $N_e=\alpha N + (1-\alpha)$, $S=N_e(1-\beta)$, and
\begin{equation}
\begin{aligned}
J_1(x) &= \frac{2}{(1-\beta)(1-e^{-2S})}\int_x^1 \frac{(e^{2S\xi}-1)(e^{-2S\xi}-e^{-2S})}{\xi(1-\xi)}d\xi \;, \\
J_2 (x) &= \frac{2}{(1-\beta)(1-e^{-2S})}\int_0^x \frac{(e^{2S\xi}-1)(1-e^{-2S\xi})}{\xi(1-\xi)}d\xi \;.
\end{aligned}
\end{equation}
\end{result}
Note that these integrals cannot be solved in closed form, so we can only estimate them numerically.

\paragraph{Numerical validation.}
We compare our approximations (\cref{eq:kimura_p,eq:kimura_t}) with results of simulations of our dichotomous model using various $\alpha$ and $\beta$ values, as well as simulations of the Wright-Fisher model, using the effective selection coefficient, $1-\beta$, and effective population size, $N_e=\alpha N + (1-\alpha)$. 
We find see that the two models have similar dynamics, and both are well approximated our approximations (\Cref{fig:var_alpha}).

\begin{figure}[h]
    \includegraphics[width=\linewidth]{../figures/final/kimura_var.pdf}
  \caption{\textbf{Fixation probability and time in a constant environment.}
  Fixation probability and time (in generations) as a function of the success-bias weight $\alpha$ (bottom x-axis), or effective population size $N_e$ (top x-axis) in the top row, and as a function of the success coefficient, $1-\beta$, on the bottom row.
  The approximation (black; \cref{eq:kimura_p}) agrees with both DM simulations (green) and Wright-Fisher simulation (orange).
  Fixation probability (A) is bounded by $2(1-\beta)$ (blue).
  Markers are averages of $10,000$ simulations, error bars show 95\% confidence intervals for panels A and B and 75\% for panels C and D.
   Here, Population size, $N=1,000$; phenotype values, $\hat{A}=1$, $A=0.7$ (A and B), $A = a \cdot \hat{A}$ with $0.01 \le a \le 0.99$ (panels C and D); success coefficient, $1-\beta=s=0.044$ (A and B); success-bias weight, $\alpha=0.01$ (panels C and D).}
  \label{fig:var_alpha}
\end{figure}

@@

Note that here, we have the average selection coefficient during a cycle of $n$ generations as the selection coefficient \cref{eq:kimura_p}.
Using the drift and diffusion terms and following \citet{changeEnv}, we can approximate the fixation probability in a changing environment using
\begin{equation}\label{eq:ch_env}
\tilde\pi(x) = \frac{1-e^{-2 \frac{S_n}{n} N_e x}}{1-e^{-2 \frac{S_n}{n} N_e}} \;.
\end{equation}

\paragraph{Numerical validation.}
% TODO SAAR edit this section
Comparing our approximation (\cref{eq:ch_env}) to simulations, we find that the approximation fits simulation results well for variable bias weights, $\alpha$, which corresponds to the effective population size (\Cref{fig:ch_env_alpha_beta}A).

@@

We found that for a large $k/l$ ratio (with a constant cycle length, $n=k+l=100$), the changing environment approximation (\cref{eq:ch_env}) converges to the constant environment approximation (\cref{eq:kimura_p}), see \Cref{fig:ch_env_alpha_beta}C and \Cref{fig:ch_env_alpha_beta}D.

The approximation follows the trend of the simulation results for all $\alpha$ values.
When increasing the success coefficient to more than 0.15, the simulation results were located above the changing environment approximation, and below the constant environment approximation. We believe the reason is the structure of the cycle.
Our proof and approximation in the changing environment are for a large amount of cycles, and when the success coefficient is too high, there might be very few cycles. Either the ideal trait is copied by enough copiers so that the influence is sufficient to negate the success bias when the cycle changes (and the trait favored by the bias becomes the disfavored), or the opposite happens, and the ideal trait gets extinct before there were enough copiers that copied it.
We then tried to change the ratio between the number of cycles where $\hat{A}$ is favored and disfavored. We showed that the approximation fits well regardless of the ratio, but when the ratio of favored generations to disfavored ones is very high, it is very similar to a constant environment model.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TODO Edit to follow our latest definition of prestige etc in the introduction
\section*{Discussion}
During cultural transmission, cultural traits such as attitudes, values, beliefs, and behavioral patterns are transmitted between individuals, for example via copying and social learning.
Some cultural traits or cultural role-models may be copied more often due to transmission biases. 
A common bias is success bias, in which copiers are more likely to copy a successful role-model. Many models assume that success can be accurately estimated.
However, it has been suggested that because it is hard to estimate success, a more common bias is \emph{prestige bias}---a bias towards role-models perceived to be successful in other traits \citep{fijian_social_bias}.
